Measuring the double-drift illusion with hand tracking

### 2021-11-24 Manuscript.Rmd
revisions

---
title: "Measuring the double-drift illusion with hand tracking"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document:
    reference_docx: template.docx
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

@@

@@

_1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\

@@

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=6, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **A:** Raw individual trajectories. In general, the traced paths approximate linearity, with the direction depending on the internal motion. **B:** Four trials with a possible reset. **C:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines show the average heading. In light blue is the average area (across participants) containing 90% of the heading data. **D:** The strength of the illusion in degrees angle deviation from the physical path. In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Black line: prediction of the vector combination model of Eq. 1 with $K=0.74$ as in Heller et al. (2021), which is indistinguishable from the best fit to our data: $K=0.738$."}
plotBoundedTracking()
```

@@

The setup was the same as used in the online tracking experiment, except that we now also used a small keypad for responses (Fig. 1, right). Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants. The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 and ~3.5 _dva/s_,

THOSE NUMBERS DON'T MAKE SENSE...

and had an internal drift of 2, 3 or 4 _cps_ .

@@

```{r fig4, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **Left:** Reset points are taken at the first sample in a splined-smoothed re-traced trajectory that goes in the direction opposite that of the expected illusion. Orange and gray lines: example trajectory. Orange dot: reset point. Red dot: point at half the distance to the reset point. Angle a: measure of illusion strength. **Middle:** One participant's trajectories with very clear 'return' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear. **Right:** Distribution of all 858 used reset points from 8/9 participants."}
#plotExampleData()
plotResetPoints()
```

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 4 (right), we can see that the reset points vary more along the y-axis, than along the x-axis. That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time. However, there is also considerable variance, so that fitting any line through the data, will necessarily leave errors.

```{r fig5, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5:** Illusion strength and reset models. **Left:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities. Purple dots are the group means from 3 s passes, orange from 4 s passes (2, 3 and 4 _cps_ internal speeds). Predictions from the vector combination model in Eq. 1: black line; $K=0.74$ as in Heller et al. (2021), light blue line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval. **Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits. Gray dots are the reset points from all participants. While a spatial limit provides the better fit, neither a spatial nor a temporal limit explains the data well."}
#plotResetPoints()
plotModels()
#plotBootstrappedFits()
#plotIllusionStrength()
```

>>>>

---
title: "Measuring the double-drift illusion with hand tracking"
author: 
- Bernard Marius 't Hart 1,*
- Sirui Liu 2
- Peter Ulrich Tse 2
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document:
    reference_docx: template.docx
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

# new addition with more classic psychophysics on perceived offsets:
source('R/exp3_gaborjumps.R')

@@

# for exp 3, we fit psychometric functions to get PSEs:
getPSEs()

@@

_1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
_2: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\
_3: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\

@@

Potential extra citation: [@PatricioDecima2022].

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=6, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **A:** Raw individual trajectories. In general, the traced paths approximate linearity, with the direction depending on the internal motion. **B:** Four trials with a possible reset. **C:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines show the average heading over time. The light blue depicts the average area (across participants) containing 90% of the heading data. The darker blue areas depict the 95% confidence interval of the mean heading. **D:** The strength of the illusion in degrees angle deviation from the physical path. In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Black line: prediction of the vector combination model of Eq. 1 with $K=0.74$ as in Heller et al. (2021), which is indistinguishable from the best fit to our data: $K=0.738$."}
plotBoundedTracking()
```

@@

The setup was the same as used in the online tracking experiment, except that we now also used a small keypad for responses (Fig. 1, right). Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants. The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 and ~3.5 _dva/s_, and had an internal drift of 2, 3 or 4 _cps_ .

@@

```{r fig4, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **A-D:** Example trajectories and reset points. All trajectory samples in transparent light blue (denser samples result in darker color). Dark red lines are, pruned, interpolated and filtered trajectories (see Methods). Purple circles indicate reset points. **A:** Trajectory with several sudden resets. **B:** 'Hit-the-wall' reset. **C:** Slow return reset. **D:** Continued accumulation of the illusion. **E:** Distribution of all 827 detected reset points. **F:** Using the angle to a point along the trajectory at half the distance to the reset point as a measure of illusion strength, we can compare this data to other data. The illusion is weaker ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
#plotExampleData()
plotResetPoints()
```

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 4E, we can see that the reset points vary more along the y-axis, than along the x-axis. This would be consistent with a spatial limit to the illusion resulting in resets. However, there is also considerable variance, so that fitting any line through the data, will necessarily leave errors.

```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models. **A:** Model limits relative to reset point locations. Reset points neither fall along a spatial limit (blue line) or a temporal limit (red lines). Marginal distributions show that reset points' X coordinates and time since stimulus approximately fall in a gamma distribution. **B:** Gamma distribution on time (log likelihood: -766.6). **C:** Gamma distribution on spatial offset (log likelihood: -1319.3)."}
#plotResetPoints()
invisible(capture.output(plotModels()))
#plotBootstrappedFits()
#plotIllusionStrength()
```
